---
title: "Simple Guidelines for Cleaning Data in R"
author: "Duncan Temple Lang and Matt Espe"
date: \today{}
geometry: margin=1in
output:
pdfdocument:
---

# Motivation

Mistakes happen. Typos, switched records, missing data - these are the
things you never have to deal with in statistics classes, yet can take
up 80\% of any data analysis effort.


```{r}
# The read.* functions import data from various formats

rice_data <- read.csv("RICE-AcreageYieldProductionandPrice-2016-05-14.csv", header = TRUE)

```

Great - now what? First, lets look at some of the basics of how the data are being stored in R.

```{r}
# class reports the class of the object
class(rice_data)

# data.frame is one of the basic data structures in R
# Although it displays similar to a Excel spreadsheet, 
# R stores it as a collection of columns

# How big is this data.frame?
dim(rice_data)
nrow(rice_data)
ncol(rice_data)

# What variables are in the data.frame?
colnames(rice_data)
str(rice_data)

# We can look at just the top and bottom of an object
head(rice_data)
tail(rice_data)

# We can pull out a single column using subsetting
# There are many different ways to subset in R
# We can pull out the column by name using the "$" operator
head(rice_data$LOCATION)

# Or we can pull out a column by number using the "[" operator
# the "[" operator subsets by index - for data.frames there are two indices
# rows and columns. The "[" works like this [row#, column#]
# if you leave one of the indices blank, R will return all values for that index

# To select the first column
head(rice_data[,1])

# And the first row
rice_data[1,]

# And finally, the value in the first row, first column
rice_data[1,1]

# Notice that each piece of the data.frame has its own class
# Many of the columns have class "factor"
# What is a factor?
# typeof will tell us the basic data type
typeof(rice_data$LOCATION)
```

- read in data

  - explain data types


# Core data types:

  - matrix vs data.frame

  - print to console vs. actual data type


# Guidelines:

  - Check class of data object

  - dimensions of data, names of variables

  - structure of the data - str()

  - summary() of the data

  - classes of variables/columns

  - <aside/teaching moment> understand the data structures

  - unique values

  - distributions (text & graphics)

    - categorical
	
    - numeric
	
    - incorrect levels (typos)
	
  - names
  
  - dims
  
  - verify random rows vs. top/bottom
  
  - missing values - distribution/location
  
    - drift
	
    - missing in time
	
    - systematic error
	
  - outliers/inliers
  
  - implied NAs (e.g. 999)
  
  - groups - check expected number in each group
  
  - merge datasets and verify the results
  
  - conversion of columns to date/time
