---
title: "Simple Guidelines for Cleaning Data in R"
author: "Matt Espe and Duncan Temple Lang"
date: \today{}
output:
pdfdocument:
---

```{r echo = FALSE}
# Set some options to make the document easier to read
options(max.print = 100)
```

# Motivation

Mistakes happen. Typos, switched records, missing data - these are the
things you never have to deal with in statistics classes, yet can take
up 80\% of any data analysis effort in the real world. In most
statistics classes, you might do the following:


```{r}
# The read.* functions import data from various formats
rice_data <- read.csv("RICE-AcreageYieldProductionandPrice-2016-05-14.csv", header = TRUE)

```

Great - now what? In classes, you are often given data that are
already cleaned and free of issues. But few will be so lucky with real
data. Where do you start?

First, lets look at some of the basics of how the data are being
stored in R. Investigating data when it is first imported into R can
help you spot issues early on.

# Basics of R objects

```{r}
# class reports the class of the object
class(rice_data)

# data.frame is one of the basic data structures in R
# Although it displays similarly to a Excel spreadsheet, 
# R stores it as a collection of columns

# How big is this data.frame?
dim(rice_data)
nrow(rice_data)
ncol(rice_data)

# What variables are in the data.frame?
colnames(rice_data)

# the str function gives us the structure of the object,
# including its class, its component pieces and their classes,
# and its dimensions.
# Very helpful to get lots of information
str(rice_data)

# We can pull out a single column using subsetting
# There are many different ways to subset in R
# We can pull out the column by name using the "$" operator
rice_data$LOCATION

# Or we can pull out a column by number using the "[" operator
# the "[" operator subsets by index - for data.frames there are two indices
# rows and columns. The "[" works like this [row#, column#]
# if you leave one of the indices blank, R will return all values for that index

# To select the first column
rice_data[,1]

# And the first row
rice_data[1,]

# And finally, the value in the first row, first column
rice_data[1,1]

# We can also look at just the top and bottom of an object
head(rice_data$LOCATION)
tail(rice_data$LOCATION)

# Using the str function, we found that each piece of the data.frame
# has its own class. Many of the columns have class "factor"
# What is a factor?
# typeof will tell us the basic data type
typeof(rice_data$LOCATION)

# This can be very confusing. When we printed the "LOCATION" column,
# it did not look like an integer, but rather names of the locations.
# Factors are a special class which are stored as integers, with each
# integer "mapped" to a level (i.e., a label)
levels(rice_data$LOCATION)

# Remember that a data.frame was a collection of columns?
# What data type is the data.frame?
typeof(rice_data)

# A data.frame is a special list where each piece has the same length.
# A list is a collection of objects, and can be very general. Each element
# of a list can be completely different classes, with different dimensions.

```

## Review

  - Inspecting the structure of your data can reveal issues early on.
  - There are fuctions in R that allow you to inspect many aspects of how data are stored and represented.
  - Subsetting allows you to isolate part of a data structure
  
# Finding problems

- read in data
  - explain data types

# Core data types:

  - matrix vs data.frame
  - print to console vs. actual data type

# Guidelines:

  - Check class of data object
  - dimensions of data, names of variables
  - structure of the data - str()
  - summary() of the data
  - classes of variables/columns
  - <aside/teaching moment> understand the data structures
  - unique values
  - distributions (text & graphics)
    - categorical
    - numeric
    - incorrect levels (typos)
  - names
  - dims
  - verify random rows vs. top/bottom  
  - missing values - distribution/location  
    - drift	
    - missing in time	
    - systematic error
  - outliers/inliers
  - implied NAs (e.g. 999) 
  - groups - check expected number in each group
  - merge datasets and verify the results
  - conversion of columns to date/time


  - Understand how data was recorded and how errors could have been introduced 
    - Dates in different countries
    - Character encodings
    - Excel thinking 
    
### Univariate Checks
+ Verify values are from appropriate set/interval
  + Look at range of each variable
  + Check levels of factors
  + See if there are spelling errors (adist())
+ Consider OpenRefine  (interactive and not necessarily reproducible)

+ Values in a row may belong to other columns  (e.g. missing a , to indicate blank cell, or too many ,)
  + Fix following columns.
  
+ Check dates make sense
  + Valid days of they year (e.g., not Feb 31st, check leap year)
  + Most dates on same day, but few not, e.g. Wednesday, and a few on Saturday.
  + Month and day possibly reversed
  + Year recorded with 4 and 2 digits.

+ Check times make sense


+ Are several observations messed up
   + If several are wrong, they may hide each other as being outliers/errors.
      + All from the same level
      + All from the same data collector/entry person
      + Repeated measurements from same 